---
title: 'Group Work Aphids'
author: "Aphid Research Team"
date: "2023-03-22"
output: 
  html_document:
    toc: yes
    theme: cerulean
    highlight: tango
---

---

## **Packages**

```{r}
library(tidyverse)   # data manipulation
library(car)         # regression
library(performance) # check model assumptions
library(ggplot2)     # data viz
library(lme4)        # mixed effects models
```

&ensp;

---

## **Research Question**


#### Question: Is aphid count influenced by the presence of birds and by season?
#### H0: Aphid count is not influenced by presence/absence of birds.
#### H0: Aphid count is not influenced by season.
#### H1: Aphid count decreases with presence of birds.
#### H2: Aphid count changes with season. 

&ensp;

---

## **Exploratory**

```{r}
# Load data
aphids <- read.table("C:\\Users\\hlear\\Desktop\\statsandviz\\R07_statistical_modeling\\data\\aphids.txt", sep = "\t", header = TRUE)


# Explore data
head(aphids)
dim(aphids)

# Visualize data

# Boxplot by treatment and season
ggplot(aphids, aes(x = Treatment, y = Counts, fill = Season)) +
  geom_boxplot() +
  labs(x = "Treatment", y = "Aphid counts", fill = "Season")+
   ggtitle("Boxplot of aphid counts by treatment and season")

# Histogram of aphid counts
ggplot(aphids, aes(x = Counts)) +
  geom_histogram() +
  labs(x = "Aphid counts", y = "Frequency")

# Scatterplot of counts by plot
ggplot(aphids, aes(x = Plot, y = Counts)) +
  geom_point() +
  labs(x = "Plot", y = "Aphid counts")

```

&ensp;

---

## **Statistical Tests**

#### ANOVA
In the case of the aphids data, we have two treatments (present or excluded) and two seasons (Season1 or Season2), which suggests that an ANOVA is appropriate. Additionally, an ANOVA can help us test for interactions between the two factors (treatment and season), which can provide more information about the effect of each factor on aphid counts.

```{r}
# Perform ANOVA
aphids.aov <- aov(Counts ~ Treatment * Season, data = aphids)
summary(aphids.aov)
```

&ensp;

---

