---
title: 'Problem Set 5: Correlation'
author: "Heatherlee Leary"
date: "2023-02-13"
output: 
  html_document:
    toc: yes
    theme: cerulean
    highlight: tango
---

---

## **Packages**

```{r}
library(tidyverse) # data wrangling and cleaning
library(ggplot2)   # data visualization
library(corrgram)  # graphical display of a correlation matrix (correlogram)
library(cowplot)   # creae publication-quality figures with ggplot2
```

&ensp;

---

## **Question 1**


In a study of hyena laughter, a researcher investigated whether sound spectral properties of hyenas’ giggles are associated with age.

&ensp;

Age(years): 2,2,2,6,9,10,13,10,14,14,12,7,11,11,14,20

Frequency (Hz): 840,670,580,470,540,660,510,520,500,480,400,650,460,500,580,500

&ensp;

a) Inspect the data using a scatterplot
b) Test the linear association between both variables
c) Assume that the data are not normally distributed, test the linear association using a non-parametric correlation coefficient

&ensp;

---

#### a) Inspect the data using a scatterplot

```{r}
# Put the data into a data frame for manipulation
hyena <- data.frame(age = c(2,2,2,6,9,10,13,10,14,14,12,7,11,11,14,20), 
                    Hz = c(840,670,580,470,540,660,510,520,500,480,400,650,460,500,580,500))

# Explore data
class(hyena)
head(hyena)
colnames(hyena)

# Quick inspection of the data
# Plot using base R
plot(hyena)
```

&ensp;

---

#### b) Test the linear association between both variables

```{r}
# Fit the linear regression model
hyena_lm <- lm(Hz ~ age, data = hyena)
summary(hyena_lm) # to see R-squared and p-value

# Pairwise matrix using Pearson's Correlation Coefficient
# Pearson's r measures the strength and direction of the association.
cor(hyena, method = "pearson")

# Hypothesis test on the correlation coefficient
cor.test(hyena$age, hyena$Hz, method = "pearson")

# Plot using ggplot2
ggplot(data=hyena, aes(x=age, y=Hz)) +
  geom_point(size=2.5, alpha=0.4) +
  geom_smooth(method="lm", se=F, size=1.5, color="firebrick4") +
  xlab("Age (years)") +
  ylab("Giggle Frequency (Hz)") +
  ggtitle("Relationship between hyena age and giggle frequency (Hz)") +
  theme_cowplot(14)
```

&ensp;

---

#### c) Test the linear association using a non-parametric correlation coefficient

```{r}
# Non-parametric rank correlation methods (Spearman, Kendall)
cor(hyena, method="spearman")
cor(hyena, method="kendall")
```

&ensp;

---

## **Session Info**

```{r}
sessionInfo()
```



